---
title: "STAT 345 Midterm Project"
font: 12pt
due: "Due April 4"
output: 
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(devtools)
devtools::install_github("abresler/nbastatR@aba9179ef644f263387c1536d6ddd26104d79cf4")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 2)

library(nbastatR)
# Collect the shot data for the Utah Jazz from 2010, 2015, 2020, and 2025
jazz_10 <- teams_shots(teams = "Utah Jazz", seasons = 2010)
jazz_15 <- teams_shots(teams = "Utah Jazz", seasons = 2015)
jazz_20 <- teams_shots(teams = "Utah Jazz", seasons = 2020)
jazz_25 <- teams_shots(teams = "Utah Jazz", seasons = 2025)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)

# Function to build a shot chart for a set of shot data and a given year
plot_shot_chart <- function(dat, year) {
  # STEP 1: Calculate the total percentage of shots made
  make_percentage <- round((sum(dat$isShotMade) / count(dat)) * 100)
  
  # STEP 2: Separate data into 10x10 bins and calculate the make percentage for the bin
  bins <- dat %>%
    mutate(xbin = round(locationX / 10) * 10, # Create x-bin by rounding locationX to the nearest 10
           ybin = round(locationY / 10) * 10) %>% # Create y-bin by rounding locationY to the nearest 10
    group_by(xbin, ybin) %>% # Group data by xbin and ybin
    summarize(total = n(), # Count the total number of shots in each bin
              made = sum(isShotMade), # Sum the shots made in each bin
              .groups = "drop") %>%
    mutate(ratio = made / total, # Calculate the percent of made shots in each bin
           more_makes = made > (total / 2)) # Boolean value to flag whether a bin has more makes than misses
  
  # STEP 3: Create the shot chart
  ggplot() +
    # Fill the bins as tiles with color representing shot frequency
    geom_tile(data = bins,
              aes(x = xbin, y = ybin, fill = total),
              width = 10, height = 10) +
    scale_fill_gradient(name = "Shot Frequency", 
                        trans = "log", # Log-transform the frequency to increase the spread among smaller frequency values for better visualization
                        low = "skyblue", high = "purple",
                        breaks = c(min(bins$total), max(bins$total)), # Only label the minimum and maximum values of the legend
                        labels = c("least", "most"),
                        guide = guide_colorbar(direction = "horizontal",
                                               title.position = "top",
                                               barwidth = 8, barheight = 1)) +
    
    # Add a dot on tiles where more shots were made than missed
    geom_point(data = bins %>% filter(more_makes), # Filter bins where more shots were made than missed
              aes(x = xbin, y = ybin, color = "More shots made than missed"),
              size = 0.2, show.legend = TRUE) +
    # Adjust the legend
    scale_color_manual(name = "", 
                       values = c("More shots made than missed" = "yellow"),
                       labels = c("More shots made\nthan missed"),
                       guide = guide_legend(override.aes = list(size = 5))) +
    
    # STEP 4: Draw the lines of the court onto the chart
    annotate("path", x = 7.5 * cos(seq(0, 2 * pi, length.out = 100)), 
             y = 7.5 * sin(seq(0, 2 * pi, length.out = 100)), color = "black") + # Hoop
    annotate("segment", x = -30, xend = 30, y = -7.5, yend = -7.5, color = "black") + # Backboard
    annotate("rect", xmin = -80, xmax = 80, ymin = -47.5, ymax = 142.5, color = "black", fill = NA) + # Outer box
    annotate("rect", xmin = -60, xmax = 60, ymin = -47.5, ymax = 142.5, color = "black", fill = NA) + # Inner box
    annotate("path", x = 60 * cos(seq(0, pi, length.out = 100)), y = 142.5 + 60 * sin(seq(0, pi, length.out = 100)), color = "black") + # Free throw arc
    annotate("path", x = 60 * cos(seq(pi, 2 * pi, length.out = 100)), y = 142.5 + 60 * sin(seq(pi, 2 * pi, length.out = 100)), color = "black", linetype = "dashed") + # Dashed free throw arc
    annotate("path", x = 40 * cos(seq(0, pi, length.out = 100)), y = 40 * sin(seq(0, pi, length.out = 100)), color = "black") + # Restricted arc
    annotate("segment", x = -220, xend = -220, y = -47.5, yend = 92.5, color = "black") + # Left three-point line
    annotate("segment", x = 220, xend = 220, y = -47.5, yend = 92.5, color = "black") + # Right three-point line
    annotate("path", x = 237.5 * cos(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), y = 237.5 * sin(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), color = "black") + # Three point arc
    annotate("path", x = 60 * cos(seq(0, pi, length.out = 100)), y = 422.5 + 60 * sin(seq(pi, 2*pi, length.out = 100)), color = "black") + # Outer center circle
    annotate("path", x = 20 * cos(seq(0, pi, length.out = 100)), y = 422.5 + 20 * sin(seq(pi, 2*pi, length.out = 100)), color = "black") + # Inner center circle
    annotate("rect", xmin = -250, xmax = 250, ymin = -47.5, ymax = 422.5, color = "black", fill = NA) + # Court boundary
    
    # STEP 5: Adjust court appearance
    coord_fixed(ratio = 1) +
    xlim(-250, 250) +
    ylim(-50, 460) +
    
    # STEP 6: Add year and make percentage labels
    labs(title = as.character(year)) + # Use the year as the plot title
    theme_void() + # Remove grid lines and the background
    theme(plot.title = element_text(face = "bold", hjust = 0.5)) + # Bold and center the year
    
    # Add the make percentage below the year
    annotate("text", x = 0, y = 457, label = paste0(as.character(make_percentage), "% made"), size = 3) 
}
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# STEP 7: Plot charts for each year
plt10 <- plot_shot_chart(jazz_10, 2010)
plt15 <- plot_shot_chart(jazz_15, 2015)
plt20 <- plot_shot_chart(jazz_20, 2020)
plt25 <- plot_shot_chart(jazz_25, 2025)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# STEP 8: Combine the plots into a single graphic
library(cowplot)
library(patchwork)

# Extract the legend from one of the plots
legend <- get_legend(plt10)

# Add the plots into one graphic with a shared legend
final <- (plt10 + plt15 + plt20 + plt25) + 
  plot_layout(guides = "collect") + # Collect the legends from all the plots
  plot_annotation(title = "Utah Jazz shots over 15 years", 
                  theme = theme(plot.title = element_text(size = 17, face = "bold", hjust = 0.5))) # Add a title

# Print the combined plot
print(final)
```

### Observations:
* Shots used to be spread more evenly across the court
* Shots are now concentrated in the three-second area and around the three-point line
* The percentage of made shots has slightly decreased

\newpage

### An more thorough analysis:
* Shots used to be spread more evenly across the court
  * The 2010 chart has less white space, showing that shots were taken all over the court.
  * The 2025 chart has more white space, meaning there were many more locations on the court where no shots were taken.
  * The 2010 chart has less color variation, maintaining a mostly blue color throughout, showing the consistency in the frequency of shots being taken throughout the court.
  * The 2025 chart has more color variation, with purple clusters surrounded by blue, showing how specific locations on the court had much higher frequencies of shots than others.
* Shots are now concentrated in the three-second area and around the three-point line
  * From chart 2010 to 2015 and 2020, the intensity of the purple color, signifying high frequency, just beyond the three-point line increases, showing the progressive increase of shots near the line.
  * From chart 2010 to 2015 and 2020, the intensity of the purple color, signifying high frequency, in the three-second area increases, showing the progressive increase of shots in the zone.
  * The 2020 and 2025 charts look very similar, suggesting that the patterns have remained fairly constant over the past several years.
* The percentage of made shots has slightly decreased
  * 49% of shots were made in 2010.
  * This decreased to 45% in 2015 and stayed lower through 2025, where 46% were made.
  * This suggests that the changing of shot locations hasn't necessarily led to a higher number of made shots.
  
```{r, echo=FALSE, warning=FALSE, message=FALSE}
jazz <- rbind(jazz_10, jazz_15, jazz_20, jazz_25)

library(ggplot2)
library(dplyr)

# Separate data into 10x10 bins and calculate make percentages
bins <- jazz %>%
  mutate(xbin = round(locationX / 10) * 10, # Create x-bin by rounding locationX to the nearest 10
         ybin = round(locationY / 10) * 10) %>% # Create y-bin by rounding locationY to the nearest 10
  group_by(xbin, ybin) %>% # Group data by xbin and ybin
  summarize(total = n(), # Count the total number of shots in each bin
            made = sum(isShotMade), # Sum the shots made in each bin
            .groups = "drop") %>%
  mutate(ratio = made / total) # Calculate the percent of made shots

# Create the shot chart
ggplot() +
  # Fill the bins as tiles with color representing shot frequency
  geom_tile(data = bins %>% filter(ratio < 0.20), # Filter bins with less than 25% made shots
            aes(x = xbin, y = ybin, fill = total),
            width = 10, height = 10) +
  scale_fill_gradient(name = "Frequency",
                      low = "pink", high = "red") +
  # Draw the lines of the court onto the chart
  annotate("path", x = 7.5 * cos(seq(0, 2 * pi, length.out = 100)), 
           y = 7.5 * sin(seq(0, 2 * pi, length.out = 100)), color = "black") + # Hoop
  annotate("segment", x = -30, xend = 30, y = -7.5, yend = -7.5, color = "black") + # Backboard
  annotate("rect", xmin = -80, xmax = 80, ymin = -47.5, ymax = 142.5, color = "black", fill = NA) + # Outer box
  annotate("rect", xmin = -60, xmax = 60, ymin = -47.5, ymax = 142.5, color = "black", fill = NA) + # Inner box
  annotate("path", x = 60 * cos(seq(0, pi, length.out = 100)), y = 142.5 + 60 * sin(seq(0, pi, length.out = 100)), color = "black") + # Free throw arc
  annotate("path", x = 60 * cos(seq(pi, 2 * pi, length.out = 100)), y = 142.5 + 60 * sin(seq(pi, 2 * pi, length.out = 100)), color = "black", linetype = "dashed") + # Dashed free throw arc
  annotate("path", x = 40 * cos(seq(0, pi, length.out = 100)), y = 40 * sin(seq(0, pi, length.out = 100)), color = "black") + # Restricted arc
  annotate("segment", x = -220, xend = -220, y = -47.5, yend = 92.5, color = "black") + # Left three-point line
  annotate("segment", x = 220, xend = 220, y = -47.5, yend = 92.5, color = "black") + # Right three-point line
  annotate("path", x = 275 * cos(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), y = 275 * sin(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), color = "green", size = 1) + # Suggested 4-point line
  annotate("path", x = 237.5 * cos(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), y = 237.5 * sin(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), color = "black") + # Three point arc
  annotate("path", x = 60 * cos(seq(0, pi, length.out = 100)), y = 422.5 + 60 * sin(seq(pi, 2*pi, length.out = 100)), color = "black") + # Outer center circle
  annotate("path", x = 20 * cos(seq(0, pi, length.out = 100)), y = 422.5 + 20 * sin(seq(pi, 2*pi, length.out = 100)), color = "black") + # Inner center circle
  annotate("rect", xmin = -250, xmax = 250, ymin = -47.5, ymax = 422.5, color = "black", fill = NA) + # Court boundary
  
  # Adjust court appearance
  coord_fixed(ratio = 1) +
  xlim(-250, 250) +
  ylim(-50, 460) +
  
  # Add a title
  labs(title = "2010-2025 Shot Locations Made Less Than 25%") + # Set title
  theme_void() + # Remove grid lines and the background
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) # Bold and center the title
```

  If the NBA were to incorporate a 4-point shot, I would suggest that the arc be a third of the distance from the 3-point arc to the center circle, similar to where the green arc is drawn on the plot. The plot shows locations on the court where, across 2010, 2015, 2020, and 2025, less than 20 percent of shots were made. All of the tiles beyond this line are pink, meaning they did not occur very frequently. I believe this line is therefore a good option for a 4-point line because it is proven to be difficult to make shots beyond it and players do not shoot from points beyond it all that often. A 4-point shot needs to be earned and shots made beyond this line would be appropriate to gift 4 points.
  A potential consequence of introducing a 4-point shot is that teams would quickly adjust to go for these shots more often. With enough training, players could improve their ability to make shots from this new line, making the frequency of made shots at this distance increase dramatically. This could lead to a 4-point shot becoming common and less exciting. A key feature of such a shot should be that they aren't made all that often, so players mastering this shot would defeat the purpose of introducing it. 