---
title: "STAT 345 Midterm Project"
font: 12pt
due: "Due April 4"
output: 
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(devtools)
devtools::install_github("abresler/nbastatR@aba9179ef644f263387c1536d6ddd26104d79cf4")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 2)

library(nbastatR)
# Collect the shot data for the Utah Jazz from 2010, 2015, 2020, and 2025
jazz_10 <- teams_shots(teams = "Utah Jazz", seasons = 2010)
jazz_15 <- teams_shots(teams = "Utah Jazz", seasons = 2015)
jazz_20 <- teams_shots(teams = "Utah Jazz", seasons = 2020)
jazz_25 <- teams_shots(teams = "Utah Jazz", seasons = 2025)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)

# Function to build a shot chart for a set of shot data and a given year
plot_shot_chart <- function(dat, year) {
  # STEP 1: Calculate the total percentage of shots made
  make_percentage <- round((sum(dat$isShotMade) / count(dat)) * 100)
  
  # STEP 2: Separate data into 10x10 bins and calculate the make percentage for the bin
  bins <- dat %>%
    mutate(xbin = round(locationX / 10) * 10, # Create x-bin by rounding locationX to the nearest 10
           ybin = round(locationY / 10) * 10) %>% # Create y-bin by rounding locationY to the nearest 10
    group_by(xbin, ybin) %>% # Group data by xbin and ybin
    summarize(total = n(), # Count the total number of shots in each bin
              made = sum(isShotMade), # Sum the shots made in each bin
              .groups = "drop") %>%
    mutate(ratio = made / total, # Calculate the percent of made shots in each bin
           more_makes = made > (total / 2)) # Boolean value to flag whether a bin has more makes than misses
  
  # STEP 3: Create the shot chart using ggplot2
  ggplot() +
    # Fill tiles on the plot with color based on frequency
    geom_tile(data = bins,
              aes(x = xbin, y = ybin, fill = total),
              width = 10, height = 10) +
    scale_fill_gradient(name = "Shot Frequency", 
                        trans = "log", # Scale the frequency by taking the log to increase the spread among smaller frequency values for better visualization
                        low = "skyblue", high = "purple",
                        breaks = c(min(bins$total), max(bins$total)), # Only label the ends of the legend
                        labels = c("least", "most"),
                        guide = guide_colorbar(direction = "horizontal",
                                               title.position = "top",
                                               barwidth = 8, barheight = 1)) +
    
    # Add a dot in tiles where more shots were made than missed
    geom_point(data = bins %>% filter(more_makes),
              aes(x = xbin, y = ybin, color = "More shots made than missed"),
              size = 0.2, show.legend = TRUE) +
    # Add a legend showing what the yellow dot symbolizes
    scale_color_manual(name = "", 
                       values = c("More shots made than missed" = "yellow"),
                       labels = c("More shots made\nthan missed"),
                       guide = guide_legend(override.aes = list(size = 5))) +
    
    # Manually draw the lines of the court onto the chart
    # Hoop
    annotate("path", x = 7.5 * cos(seq(0, 2 * pi, length.out = 100)), 
             y = 7.5 * sin(seq(0, 2 * pi, length.out = 100)), color = "black") +
    # Backboard
    annotate("segment", x = -30, xend = 30, y = -7.5, yend = -7.5, color = "black") +
    # Outer box (paint)
    annotate("rect", xmin = -80, xmax = 80, ymin = -47.5, ymax = 142.5, color = "black", fill = NA) +
    # Inner box (paint)
    annotate("rect", xmin = -60, xmax = 60, ymin = -47.5, ymax = 142.5, color = "black", fill = NA) +
    # Free throw arcs
    annotate("path", x = 60 * cos(seq(0, pi, length.out = 100)), y = 142.5 + 60 * sin(seq(0, pi, length.out = 100)), color = "black") +
    annotate("path", x = 60 * cos(seq(pi, 2 * pi, length.out = 100)), y = 142.5 + 60 * sin(seq(pi, 2 * pi, length.out = 100)), color = "black", linetype = "dashed") +
    # Restricted arc
    annotate("path", x = 40 * cos(seq(0, pi, length.out = 100)), y = 40 * sin(seq(0, pi, length.out = 100)), color = "black") +
    # Three-point line
    annotate("segment", x = -220, xend = -220, y = -47.5, yend = 92.5, color = "black") +
    annotate("segment", x = 220, xend = 220, y = -47.5, yend = 92.5, color = "black") +
    annotate("path", x = 237.5 * cos(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), y = 237.5 * sin(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), color = "black") +
    # Center circle
    annotate("path", x = 60 * cos(seq(0, pi, length.out = 100)), y = 422.5 + 60 * sin(seq(pi, 2*pi, length.out = 100)), color = "black") +
    annotate("path", x = 20 * cos(seq(0, pi, length.out = 100)), y = 422.5 + 20 * sin(seq(pi, 2*pi, length.out = 100)), color = "black") +
    # Court boundary lines
    annotate("rect", xmin = -250, xmax = 250, ymin = -47.5, ymax = 422.5, color = "black", fill = NA) +
    
    # Set court limits to accurately reflect dimensions
    coord_fixed(ratio = 1) +
    xlim(-250, 250) +
    ylim(-50, 460) +
    
    # Add the year above the plot
    labs(title = as.character(year)) +
    
    # Clear the background of the chart and any existing excess features
    theme_void() +
    
    # Center the year
    theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
    
    # Add the make percentage below the year
    annotate("text", x = 0, y = 457, label = paste0(as.character(make_percentage), "% made"), size = 3) 
}
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Make a plot for each of the years
plt10 <- plot_shot_chart(jazz_10, 2010)
plt15 <- plot_shot_chart(jazz_15, 2015)
plt20 <- plot_shot_chart(jazz_20, 2020)
plt25 <- plot_shot_chart(jazz_25, 2025)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(cowplot)
library(patchwork)

# Extract the legend from one of the plots
legend <- get_legend(plt10)

# Add the plots into one graphic with a single legend
final <- (plt10 + plt15 + plt20 + plt25) + 
  plot_layout(guides = "collect") + 
  plot_annotation(title = "Utah Jazz shots over 15 years", 
                  theme = theme(plot.title = element_text(size = 17, face = "bold", hjust = 0.5)))

print(final)
```

### Observations:
* Shots used to be spread more evenly across the court
* Shots are now concentrated in the three-second area and around the three-point line
* The percentage of made shots has slightly decreased

\newpage

### An more thorough analysis:
* Shots used to be spread more evenly across the court
  * The 2010 chart has less white space, showing that shots were taken all over the court.
  * The 2025 chart has more white space, meaning there were many more locations on the court where no shots were taken.
  * The 2010 chart has less color variation, maintaining a mostly blue color throughout, showing the consistency in the frequency of shots being taken throughout the court.
  * The 2025 chart has more color variation, with purple clusters surrounded by blue, showing how specific locations on the court had much higher frequencies of shots than others.
* Shots are now concentrated in the three-second area and around the three-point line
  * From chart 2010 to 2015 and 2020, the intensity of the purple color, signifying high frequency, just beyond the three-point line increases, showing the progressive increase of shots near the line.
  * From chart 2010 to 2015 and 2020, the intensity of the purple color, signifying high frequency, in the three-second area increases, showing the progressive increase of shots in the zone.
  * The 2020 and 2025 charts look very similar, suggesting that the patterns have remained fairly constant over the past several years.
* The percentage of made shots has slightly decreased
  * 49% of shots were made in 2010.
  * This decreased to 45% in 2015 and stayed lower through 2025, where 46% were made.
  * This suggests that the changing of shot locations hasn't necessarily led to a higher number of made shots.
  
  
