---
title: "4-Point Arc Proposal"
author: "Lucy Warren"
date: "2025-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 2)

library(nbastatR)
# Collect the shot data for the Utah Jazz from 2010, 2015, 2020, and 2025
jazz_10 <- teams_shots(teams = "Utah Jazz", seasons = 2010)
jazz_15 <- teams_shots(teams = "Utah Jazz", seasons = 2015)
jazz_20 <- teams_shots(teams = "Utah Jazz", seasons = 2020)
jazz_25 <- teams_shots(teams = "Utah Jazz", seasons = 2025)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
jazz <- rbind(jazz_10, jazz_15, jazz_20, jazz_25)

library(ggplot2)
library(dplyr)

# Separate data into 10x10 bins and calculate make percentages
bins <- jazz %>%
  mutate(xbin = round(locationX / 10) * 10, # Create x-bin by rounding locationX to the nearest 10
         ybin = round(locationY / 10) * 10) %>% # Create y-bin by rounding locationY to the nearest 10
  group_by(xbin, ybin) %>% # Group data by xbin and ybin
  summarize(total = n(), # Count the total number of shots in each bin
            made = sum(isShotMade), # Sum the shots made in each bin
            .groups = "drop") %>%
  mutate(ratio = made / total) # Calculate the percent of made shots

# Create the shot chart
ggplot() +
  # Fill the bins as tiles with color representing shot frequency
  geom_tile(data = bins %>% filter(ratio < 0.20), # Filter bins with less than 25% made shots
            aes(x = xbin, y = ybin, fill = total),
            width = 10, height = 10) +
  scale_fill_gradient(name = "Frequency",
                      low = "pink", high = "red") +
  # Draw the lines of the court onto the chart
  annotate("path", x = 7.5 * cos(seq(0, 2 * pi, length.out = 100)), 
           y = 7.5 * sin(seq(0, 2 * pi, length.out = 100)), color = "black") + # Hoop
  annotate("segment", x = -30, xend = 30, y = -7.5, yend = -7.5, color = "black") + # Backboard
  annotate("rect", xmin = -80, xmax = 80, ymin = -47.5, ymax = 142.5, color = "black", fill = NA) + # Outer box
  annotate("rect", xmin = -60, xmax = 60, ymin = -47.5, ymax = 142.5, color = "black", fill = NA) + # Inner box
  annotate("path", x = 60 * cos(seq(0, pi, length.out = 100)), y = 142.5 + 60 * sin(seq(0, pi, length.out = 100)), color = "black") + # Free throw arc
  annotate("path", x = 60 * cos(seq(pi, 2 * pi, length.out = 100)), y = 142.5 + 60 * sin(seq(pi, 2 * pi, length.out = 100)), color = "black", linetype = "dashed") + # Dashed free throw arc
  annotate("path", x = 40 * cos(seq(0, pi, length.out = 100)), y = 40 * sin(seq(0, pi, length.out = 100)), color = "black") + # Restricted arc
  annotate("segment", x = -220, xend = -220, y = -47.5, yend = 92.5, color = "black") + # Left three-point line
  annotate("segment", x = 220, xend = 220, y = -47.5, yend = 92.5, color = "black") + # Right three-point line
  annotate("path", x = 275 * cos(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), y = 275 * sin(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), color = "green", size = 1) + # Suggested 4-point line
  annotate("path", x = 237.5 * cos(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), y = 237.5 * sin(seq(22 * pi / 180, 158 * pi / 180, length.out = 100)), color = "black") + # Three point arc
  annotate("path", x = 60 * cos(seq(0, pi, length.out = 100)), y = 422.5 + 60 * sin(seq(pi, 2*pi, length.out = 100)), color = "black") + # Outer center circle
  annotate("path", x = 20 * cos(seq(0, pi, length.out = 100)), y = 422.5 + 20 * sin(seq(pi, 2*pi, length.out = 100)), color = "black") + # Inner center circle
  annotate("rect", xmin = -250, xmax = 250, ymin = -47.5, ymax = 422.5, color = "black", fill = NA) + # Court boundary
  
  # Adjust court appearance
  coord_fixed(ratio = 1) +
  xlim(-250, 250) +
  ylim(-50, 460) +
  
  # Add a title
  labs(title = "2010-2025 Shot Locations Made Less Than 25%") + # Set title
  theme_void() + # Remove grid lines and the background
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) # Bold and center the title
```

  If the NBA were to incorporate a 4-point shot, I would suggest that the arc be a third of the distance from the 3-point arc to the center circle, similar to where the green arc is drawn on the plot. The plot shows locations on the court where, across 2010, 2015, 2020, and 2025, less than 20 percent of shots were made. All of the tiles beyond this line are pink, meaning they did not occur very frequently. I believe this line is therefore a good option for a 4-point line because it is proven to be difficult to make shots beyond it and players do not shoot from points beyond it all that often. A 4-point shot needs to be earned and shots made beyond this line would be appropriate to gift 4 points.
  A potential consequence of introducing a 4-point shot is that teams would quickly adjust to go for these shots more often. With enough training, players could improve their ability to make shots from this new line, making the frequency of made shots at this distance increase dramatically. This could lead to a 4-point shot becoming common and less exciting. A key feature of such a shot should be that they aren't made all that often, so players mastering this shot would defeat the purpose of introducing it. 
